CODECHECK: An open-science initiative for the independent execution of computations underlying research articles.

### 2021-01-11 codecheck-paper.tex
update dimensions figure and corresponding text

The main contribution of this work is a thorough description of a process
and its variations to integrate a much needed evaluation of the 
computational reproducibility \cite{barba_terminologies_2018} into peer
review, and a demonstration of its feasibility by means of over 20
reproductions across different scientific disciplines.
We call this system CODECHECK as outlined next.

@@

\textbf{Daniel to fix by merging turnaround time into other
  dimensions, and labelling these as the four dimensions: When?
  Openness? Who? Importance?}

@@

\subsubsection*{When to do a CODECHECK and with what power}
\label{when-to-do-a-codecheck}

The point in time when a CODECHECK happens greatly depends on the 
intentions and the ascribed
importance. The earlier a CODECHECK happens in the process, the more the
results can play a role in the most important decision in scholarly 
communication: is a paper published, sent back for revisions, or rejected.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.
% TODO add a sentence about the possible effect of just asking for
% code/data early on in the submission process, see https://twitter.com/tsuyomiyakawa/status/1230672758163402752?s=09)

A pre-review CODECHECK allows editors to use it as a filter to only send 
a submission out to peer review if it passes the check. A less strict
pre-review check would simply be included in the submission package provided
to the reviewers, so they can include the CODECHECK's results in their 
evaluation of the work.
A CODECHECK may also be conducted in parallel to the scientific review. This
puts less burden on the turnaround time for the check yet only makes the 
outcomes available for the final consideration of the handling editor.
The check could also be tasked on demand, e.g., only if one of the reviewers
suggests/requests one because previous screenings might not have caught the 
fact that a computational workflow is part of the submission. However, 
soliciting such a "specialist review" is much more undesirable than having
a regular CODECHECK, without any special treatment for some submissions.
A CODECHECK could then be awarded equal weight as the other reviews.
A post acceptance CODECHECK puts the least requirements on the author and may
simply lead to an award of excellence on top of the acceptance of the
submission. This is the least strict solution while still properly 
acknowledging the check results, because the CODECHECK is to be completed 
before the publication of the final paper.
The GIScience group of checks (see \nameref{register}) falls into this
category: the AGILE conference highlights articles where the reproducibility
was successfully reviewed with a badge on the volume and article landing
pages.

Finally, a CODECHECK may also be conducted after publication, though it
would required to update the article later with a note to reference the check,
otherwise readers would presumably not be aware of it. While this has the
least impact on the current publishing practices, it also has the smallest
effect and does not acknowledge the importance of reproducible workflows for
ensuring good scientific practice.

@@

The options for the person who can conduct a CODECHECK require a not quite
simple matching of a combination of skills and availability. Ideally, the
codechecker has a matching code \emph{and} domain expertise, for example to
check a workflow based on Python code analysing neuroimaging datasets. However,
a well-documented workflow should be executable by any person with a basic
understanding of computers. Naturally, the more prerequisite knowledge the
codechecker has, the less time is needed to understand the goals and 
the mechanics of an analysis. From our experiences, the priority should be
given to matching technical expertise first, as the a lack of knowledge in
setting  up a computing environment with a particular language or tool is 
much more discouraging and the assessment of the outcome, e.g., comparing
created figures with the original, is largely possible without 
understanding.
The time and expertise allotted to the check
are the main drivers for the depth of checking, though in general, we 
expect a CODECHECK not to evaluate performance, e.g., does a racing car
have the best apex speed, but to assess operation free of failure, i.e., 
is it a car with 4 wheels and at least one door, and does it move forwards
when accelerating. This depth can be seen independent from the power of
the check (see above), as long as it is communicated clearly to all
stakeholders, most importantly authors and readers.

The concrete people to provide these expertises can range from researchers,
possibly drawn from the regular pool of volunteer reviewers or from a 
special group, to editorial staff of a publisher. If the regular reviewers
are relied upon, the editor's task to find suitable ones becomes more
tedious; extra information such as software familiarity would have to be
modelled in review platforms. Keeping diverse specialists as staff to
realise a codechecking process is unlikely to be feasible, especially for 
small or independent publishers.

We consider a single codechecker to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, should be a very factual process and not
require much debate or opinion. Depending on the power of the CODECHECK
(see TODO: check ref \ref{when-to-do-a-codecheck}), a mediator, e.g., the editor, or a
second opinion might be warranted in rare cases.

We see a great opportunity to especially involve early-career researchers
(ECRs) as codecheckers. They arguably have a high interest in learning
about the latest tools and technologies as they are building up their own
expertise and specialisation. The possibility to gain insights into latest
science for ECRs and at the same time increase the appreciation of
reproduction efforts is a sweet spot that CODECHECK shares
with \emph{ReScience X}, a journal devoted to reproduction and replication 
experiments \cite{roesch_new_2020}.
In fact, ECRs are often much more familiar with latest technology and are 
likely to be early adopters as creators, i.e., be authors of CODECHECK-ready manuscripts, too\footnote{Cf. the 77\% of 141 registered reports who were 
submitted to leading Neuroscience journals by ECRs
\cite{chambers_registered_2019}.}
They are also introduced to peer review and 
may transition into the role of scientific reviewer over time.
Not to solve the general problems of peer review, which is largely an
unsupervised process, i.e., rarely are reviewers taught how to do it, but
a junior codechecker may be supported by a senior codechecker, too.
We see a high potential in setting up the new processes for CODECHECKING with
a clear commitment on openness and transparency, independent of the
respective current peer review process (see \nameref{who-knows-who}).

@@

\subsubsection*{Who knows who?}\label{who-knows-who}

\subsubsection*{How is the CODECHECK integrated in other processes?}\label{turnaround-time}

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.},
% mimosa is another platform idea, but no real prototype AFAICT:
% https://projects.invisionapp.com/share/E9Z3RKE7W3P#/screens/436978924
% https://docs.google.com/presentation/d/1V_K8hghgnvGfEtW7TdwtowTveZyy4Qs8rMemNyHx_Dg/edit#slide=id.p
there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
The efforts include considerable time to communicate about the process, 
especially around who publishes what and when, so that proper 
cross-referencing between paper and certificate is ensured.
When integrated into peer review systems, this handling of the steps should
become much more effective.

>>>>

The main contribution of this work is a thorough description of a process
and its variations to integrate a much needed evaluation of the 
computational reproducibility \cite{barba_terminologies_2018} into peer
review, and a demonstration of its feasibility by means of 25
reproductions across different scientific disciplines.
We call this system CODECHECK as outlined next.

@@

@@

\subsubsection*{When to do a CODECHECK and with what importance?}
\label{when-to-do-a-codecheck}

The point in time when a CODECHECK happens greatly depends on the 
intentions and the ascribed importance.
The earlier a CODECHECK happens in the process, the more the
results can play a role in editorial decisions:
is a paper published, sent back for revisions, or rejected.
Such a \textbf{strict prerequisite} can reduce the workload of reviewers,
and therefore seems out of place late in the review process.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.
% TODO add a sentence about the possible effect of just asking for
% code/data early on in the submission process, see https://twitter.com/tsuyomiyakawa/status/1230672758163402752?s=09)

We see a CODECHECK as part of a typical scholarly communication process.
While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.},
% mimosa is another platform idea, but no real prototype AFAICT:
% https://projects.invisionapp.com/share/E9Z3RKE7W3P#/screens/436978924
% https://docs.google.com/presentation/d/1V_K8hghgnvGfEtW7TdwtowTveZyy4Qs8rMemNyHx_Dg/edit#slide=id.p
there is a value in an evolutionary approach.
Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
These efforts include considerable time to communicate about the process, 
especially around who publishes which document when, so that proper 
cross-referencing between paper and certificate is ensured via persistent
identifiers.
When integrated into a peer review system, this handling of documents should
become much more effective and reduce organisational efforts.

A \textbf{pre-review} CODECHECK allows editors to use it as a filter to only
send  a submission out to peer review if it passes the check.
The certificate would  be included in the submission package provided
to the reviewers, so they can include the CODECHECK's results in their 
evaluation of the work.
Reviewers may then judge the the relevance of the computations for the results
of the work, or follow journal guidelines as how strict to apply the check's
outcomes.

A CODECHECK may also be conducted in \textbf{parallel} to the scientific review.
This puts less burden on the turnaround time for the check, yet only makes the 
outcomes available for the final consideration of the handling editor.
The check could also be tasked on demand, e.g., only if one of the reviewers
suggests/requests one because previous screenings might not have caught the 
fact that a computational workflow is part of the submission. However, 
soliciting such a "specialist review" is much more undesirable than having
a regular CODECHECK, without any special treatment for some submissions.
A CODECHECK could then be awarded \textbf{equal weight} as the other reviews.

A \textbf{post acceptance} CODECHECK puts the least requirements on the author 
and may simply lead to an award of excellence on top of the acceptance of the
submission. This \textbf{extra merit} is the least strict solution while still
properly  acknowledging the check results,
because the CODECHECK is to be completed 
before the publication of the final paper.
The GIScience group of checks (see \nameref{register}) falls into this
category: the AGILE conference highlights articles where the reproducibility
was successfully reviewed with a badge on the volume and article landing
pages. Similarly, the GIScience papers in collaborations with journals were 
checked while the authors worked on revisions.

Finally, a CODECHECK may also be conducted \textbf{post publication}, though it
would required to update the article and article metadata later to reference 
the check, otherwise readers would presumably not be aware of it. In general,
publishers are hesitant to make such a revision of published articles.
While this has the least impact on the current publishing practices and also
enables open discourse,
it also has the smallest effect on the original submission
and does not acknowledge the importance of reproducible workflows for
ensuring good scientific practice.

\subsubsection*{Openness, or: Who knows who?}\label{who-knows-who}

@@

The options for the person who can conduct a CODECHECK require a not quite
simple matching of a combination of skills and availability. Ideally, the
codechecker has a matching code \emph{and} domain expertise, for example to
check a workflow based on Python code analysing neuroimaging datasets. However,
a well-documented workflow should be executable by any person with a basic
understanding of computers. Naturally, the more prerequisite knowledge the
codechecker has, the less time is needed to understand the goals and 
the mechanics of an analysis. From our experiences, the priority should be
given to matching technical expertise first, as the a lack of knowledge in
setting  up a computing environment with a particular language or tool is 
much more discouraging and the assessment of the outcome, e.g., comparing
created figures with the original, is largely possible without 
understanding the science.
The time and expertise allotted to the check
are the main drivers for the depth of checking, though in general, we 
expect a CODECHECK not to evaluate performance, e.g., does a racing car
have the best apex speed, but to assess operation free of failure, i.e., 
is it a car with 4 wheels and at least one door, and does it move forwards
when accelerating. This depth can be seen independent from the power of
the check (see above), as long as it is communicated clearly to all
stakeholders, most importantly authors and readers.

The concrete people to provide these expertises can range from researchers,
possibly drawn from the regular pool of volunteer
\textbf{scientific reviewers} or from a special group of
\textbf{reproducibility reviewers}, via specific roles
such as \textbf{reproducibility editors}, to experts working as part of
\textbf{editorial staff} with a publisher. If the regular reviewers
are relied upon, the editor's task to find suitable ones becomes more
tedious; extra information such as software familiarity would have to be
modelled in review platforms.

We consider a \emph{single codechecker} to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, actually match should be a very factual 
process and not
(see TODO: check ref \ref{when-to-do-a-codecheck}), a mediator, e.g., the editor, or a
second opinion might be warranted in rare cases.

We see a great opportunity to especially involve
\emph{early-career researchers}
(ECRs) as codecheckers. They arguably have a high interest in learning
about the latest tools and technologies as they are building up their own
expertise and specialisation. The possibility to gain insights into latest
science for ECRs and at the same time increase the appreciation of
reproduction efforts is a sweet spot that CODECHECK shares
with \emph{ReScience X}, a journal devoted to reproduction and replication 
experiments \cite{roesch_new_2020}.
In fact, ECRs are often much more familiar with latest technology and are 
likely to be early adopters as creators, i.e., be authors of CODECHECK-ready manuscripts, too\footnote{Cf. the 77\% of 141 registered reports who were 
submitted to leading Neuroscience journals by ECRs
\cite{chambers_registered_2019}.}
They are also introduced to peer review and 
may transition into the role of scientific reviewer over time.
Not to solve the general problems of peer review, which is largely an
unsupervised process, i.e., rarely are reviewers taught how to do it, but
a junior codechecker may be supported by a senior codechecker, too.
We see a high potential in setting up the new processes for CODECHECKING with
a clear commitment on openness and transparency, independent of the
respective current peer review process (see \nameref{who-knows-who}).

@@
